---
title: "607 Tidyverse Lubridate Extend Assignment"
author: "Ariann Chai and Stephen Phillips"
date: "2023-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lubridate

The tidyrverse package I used was lubridate which is used for date formatting and math.

```{r libraries, warning=FALSE}
library(tidyr)
library(lubridate)
library(readr)
library(tidyverse)
```

## Data

The dataset I used was "UK National Lottery - results 1994-2023" by Delta Romeo off of Kaggle (https://www.kaggle.com/datasets/deltaromeo/uk-national-lottery-results-1994-2023/).  
This dataset had four columns depicted to the date (weekday, day, month, year). My goal was to condense these columns which I ended leaving as two: weekday and date. Weekdays was edited to have the full name of weekday and date now contains the formatted year, month, and day.  

```{r intial data}
lotteryInitial <- read.csv("https://raw.githubusercontent.com/sphill12/DATA607/main/Lottery%20Order.csv")
lotteryInitial = lotteryInitial[-1,]
head(lotteryInitial)
colnames(lotteryInitial)
```

```{r clean data}
lotteryClean <- subset(lotteryInitial, select=-c(4,5))
colnames(lotteryClean)[3] = "Date"
head(lotteryClean)

for (i in 1:length(lotteryClean)){
  #get for lotteryInitial - DD MMM YYYY
  day = lotteryInitial$DD
  month = lotteryInitial$MMM
  year = lotteryInitial$YYYY
  date =  paste(year, month, day, sep="")
  lotteryClean$Date = ymd(date)
  lotteryClean$Day = wday(ymd(date),label = TRUE, abbr = FALSE)
}

head(lotteryClean)
```

## Extension

### It is useful to be able to concatenate day/week/year into one date, but sometimes the opposite is needed
```{r}
lotteryClean <- lotteryClean %>% mutate(year = year(Date))

```
### The code above extracts the year from the date. Similarly, you can use the month() and day() function to get the month and day.
```{r}
head(lotteryClean)
```
### Sometimes it is not useful to have a specific date, and it is desireable for it to be rounded down

```{r}
lotteryClean <- lotteryClean %>% mutate(date_floored = floor_date(Date, unit = "month"))
```
```{r}
head(lotteryClean)
```
### Sometimes we would like for it to be rounded up

```{r}
lotteryClean <- lotteryClean %>% mutate(date_ceiling = ceiling_date(Date, unit = "month"))
```

```{r}
head(lotteryClean)
```
### Sometimes we want to know if a date falls within a certain timeframe. This can be done with the interval function, and the %within% function 
```{r}
date_interval <- interval(ymd("2023-9-01"), ymd("2023-11-01"))

```
```{r}
lotteryClean$Date[1] %within% date_interval
```
```{r}
lotteryClean$Date[500] %within% date_interval
```

### These are some useful functions, in addition to what Ariann posted. Thanks!


