---
title: "607 Tidyverse Lubridate"
author: "Ariann Chai"
date: "2023-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lubridate

The tidyrverse package I used was lubridate which is used for date formatting and math.

```{r libraries, warning=FALSE}
library(tidyr)
library(lubridate)
```

## Data

The dataset I used was "UK National Lottery - results 1994-2023" by Delta Romeo off of Kaggle (https://www.kaggle.com/datasets/deltaromeo/uk-national-lottery-results-1994-2023/).  
This dataset had four columns depicted to the date (weekday, day, month, year). My goal was to condense these columns which I ended leaving as two: weekday and date. Weekdays was edited to have the full name of weekday and date now contains the formatted year, month, and day.  

```{r intial data}
lotteryInitial <- read.csv("lotteryorder.csv")
lotteryInitial = lotteryInitial[-1,]
head(lotteryInitial)
colnames(lotteryInitial)
```

```{r clean data}
lotteryClean <- subset(lotteryInitial, select=-c(4,5))
colnames(lotteryClean)[3] = "Date"
head(lotteryClean)

for (i in 1:length(lotteryClean)){
  #get for lotteryInitial - DD MMM YYYY
  day = lotteryInitial$DD
  month = lotteryInitial$MMM
  year = lotteryInitial$YYYY
  date =  paste(year, month, day, sep="")
  lotteryClean$Date = ymd(date)
  lotteryClean$Day = wday(ymd(date),label = TRUE, abbr = FALSE)
}

head(lotteryClean)
```
