---
title: "TidyVerse Recipe"
author: "Zainab.O"
date: "2023-10-29"
output: html_document
---
OKAY , so I am being asked to CREATE a project using tidyverse


Let's get into the assignment 

The purpose of this assignment is the following:
Your task here is to Create an Example.  Using one or more TidyVerse packages, and any dataset from fivethirtyeight.com or Kaggle, create a programming sample “vignette” that demonstrates how to use one or more of the capabilities of the selected TidyVerse package with your selected data set.

We decided on tattoo data because I like them but my mother and Co believe them to be things for the heathens
so the article touch me and my stressed out heart.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


DOWNLOAD PACKAGES
```{r}
library(tidyverse)
library(dplyr)
library(readr)
```

```{r}
datadf = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/nba-tattoos/nba-tattoos-data.csv")
```
```{r}
head(datadf)
```
```{r}
summary(datadf)
```
#Here I used  pipe and the group_by_all function to count all the "no" and "yes"  and create a new dataframe
```{r}
newdf = datadf %>%
  group_by(`Tattoos yes/no`) %>%
  count()

newdf
```
#Here I have a bar graph demonstrating the frequency of no tattoos vs tattoos of the basketball players in my data
```{r}
newdf %>%
  ggplot(aes(x=`Tattoos yes/no`,y=n, fill=`Tattoos yes/no`))+
  geom_bar(stat = "identity")+
  ggtitle("The Count of Basketball Players with Tattos vs those without")
```

This bar graph demonstrates that there is  a very small difference in count between basketball players that have tattoos versus those that do not have tattoo. I made this graph using  ggplot to create a bar graph and tidyverse to group and count all the "no" together and counted all the "yes" together. 
```{r cars}

```


Reference 
data : https://raw.githubusercontent.com/fivethirtyeight/data/master/nba-tattoos/nba-tattoos-data.csv
Article: https://fivethirtyeight.com/features/what-ethan-swan-learned-from-tracking-every-tattoo-in-the-nba/

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
