---
title: "tidyverse: scaling and labeling for ggplots"
author: "Kelly Eng"
date: "2023-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions for ggplot2

## Comparing Avengers

Source: https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/

According to the article, 69 had died at least once.

```{r, warning=FALSE}

# Read csv file from fivethirtyeight.com's GitHub and use the pipe operator to select only the columns we want
avengers <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv") %>%
  filter(!is.na(`Name/Alias`)) %>%
  select(`Name/Alias`, Appearances, `Current?`, Gender, Year, `Years since joining`, Honorary, Death1, Death2, Death3, Death4, Death5)

# Gets all the indices that are null for Death2
indices_for_na_death2 <- is.na(avengers$Death2)
# Fill the NAs with the NO string
avengers$Death2[indices_for_na_death2] = "NO"

# Repeat for 3 or more deaths with NA values
indices_for_na_death3 <- is.na(avengers$Death3)
avengers$Death3[indices_for_na_death3] = "NO"
indices_for_na_death4 <- is.na(avengers$Death4)
avengers$Death4[indices_for_na_death4] = "NO"
indices_for_na_death5 <- is.na(avengers$Death5)
avengers$Death5[indices_for_na_death5] = "NO"

# There is a large gap for the heroes that debuted in 1900 and those that did later, the year closest to 1900 is 1963
avengers <- avengers %>%
  filter(Year != 1900)

# Create new column using the mutate function and case_when
# case_when comes from dplyr in tidyverse, it checks multiple cases compared to the binary results of ifelse
# Default condition is TRUE, in this case if there are no deaths, that row should be given a value of 0
avengers <- avengers %>%
  mutate(num_deaths = case_when(
    Death5 == "YES" ~ "5",
    Death4 == "YES" ~ "4",
    Death3 == "YES" ~ "3",
    Death2 == "YES" ~ "2",
    Death1 == "YES" ~ "1",
    TRUE ~ "0"
  ))
```

## 1A: basic ggplot

Q: How much appearances do the Avengers have since their debut?
A: Use ggplot to compare

```{r}
avengers %>%
  ggplot(aes(x = Year, y = Appearances, color = num_deaths)) +
  labs(x = "Debut Year", y = "Total Appearances", title = "Avengers") +
  geom_point()
```
## 1B How do I add labels?

```{r}
avengers %>%
  ggplot(aes(x = Year, y = Appearances, color = num_deaths)) +
  labs(x = "Debut Year", y = "Total Appearances", title = "Avengers") +
  geom_text(aes(label = num_deaths)) +
  geom_point() 
```

## 2A Adding Scaling

Q: How do we scale the y-axis in log10 scale to make the points less clustered together?
A: Use  scale_y_log10() 

```{r}
avengers %>%
  ggplot(aes(x = Year, y = Appearances, color = num_deaths)) +
  labs(x = "Debut Year", y = "Total Appearances", title = "Avengers") +
  geom_text(aes(label = num_deaths)) +
  scale_y_log10() +
  geom_point() 
```

## 2B Moving Labels on the X-axis

Q: How do we move labels horizontally?
A: Use the nudge_x argument in geom_text(), use positive numbers to move the labels right and negative to move them left

```{r}
avengers %>%
  ggplot(aes(x = Year, y = Appearances, color = num_deaths)) +
  labs(x = "Debut Year", y = "Total Appearances", title = "Avengers") +
  geom_text(aes(label = num_deaths), nudge_x = -0.7, size = 3) +
  scale_y_log10() +
  geom_point() 
```

## 2C Moving Labels on the Y-axis

Q: How do we move labels vertically?
A: Use the nudge_y argument in geom_text(), use positive numbers to move the labels up and negative to move them down

```{r}
avengers %>%
  ggplot(aes(x = Year, y = Appearances, color = num_deaths)) +
  labs(x = "Debut Year", y = "Total Appearances", title = "Avengers") +
  geom_text(aes(label = num_deaths), nudge_x = -0.7, nudge_y = 0.1, size = 3) +
  scale_y_log10() +
  geom_point() 
```
