---
title: 'tidyverse: using forcats to improve your ggplots'
author: "catlin"
date: "2/10/2021"
output:
  pdf_document: default
  html_document: default
---

Note that although forcats is part of the "tidyverse", it is not automatically loaded when you run `library(tidyverse)`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(forcats)
```

# Handy forcats functions for ggplot2

## Comparing followers of world religions

Source: https://en.wikipedia.org/wiki/List_of_religious_populations

I was looking for a simple dataset with count data for many items to demonstrate some basic forcats functions that are useful when creating plots.

```{r, warning=FALSE}
religions = read_csv("https://raw.githubusercontent.com/acatlin/data/master/religions.csv", 
                  col_names = FALSE) %>% 
    rename(religion = X1, followers = X2) %>% 
    mutate(millions_of_followers = followers/1000000.0) %>% 
  select(religion, millions_of_followers)

religions
```

## 1A: basic ggplot

Q: What are the most followed religions?
A: Use ggplot to compare religious populations

You can also embed plots, for example:

```{r}
religions %>% 
  ggplot(aes(x = religion, y = millions_of_followers)) + 
    geom_col(fill = "lightblue") + 
    labs(x = "religion", y = "millions of followers", 
           caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") 
```
 
## 1B: How do I flip coordinates?

```{r}
religions %>% 
  ggplot(aes(x = religion, y = millions_of_followers)) + 
    geom_col(fill = "lightblue") + 
    labs(x = "religion", y = "millions of followers", 
           caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") + coord_flip()
```

## 2A: How do I change sort order?
Revised by: Andy Catlin

Q: How do we change the chart to show the most followed religions first?
A: Use forcats::fct_reorder()

```{r}
library(forcats) 

ggplot(religions, aes(x = fct_reorder(religion, millions_of_followers), 
                      y = millions_of_followers)) + 
  geom_col(fill = "lightblue") + 
  labs(x = "religion", y = "millions of followers", 
      caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") + 
  coord_flip()
```

## 2B: How do I combine less frequently used categories?

Q: How do we combine the less-followed religions into a single group?
A: Use forcats::fct_other()

```{r}
top5 = unlist(select(head(arrange(religions, desc(millions_of_followers)), 5), religion))

religions %>% 
  mutate(religion = fct_other(religion, keep = top5, other_level = "Other religions")) %>%     
  ggplot(aes(x = fct_reorder(religion, millions_of_followers), y = millions_of_followers)) +
    geom_col(fill = "lightblue") + 
    labs(x = "religion", y = "millions of followers", 
        caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") + 
        coord_flip()
```

## 2C: Adding a title
Reference: https://www.geeksforgeeks.org/ggplot2-title-and-subtitle-with-different-size-and-color-in-r/ 

```{r}
religions %>% 
  mutate(religion = fct_other(religion, keep = top5, other_level = "Other religions")) %>%     
  ggplot(aes(x = fct_reorder(religion, millions_of_followers), y = millions_of_followers)) +
    geom_col(fill = "lightblue") + 
    labs(x = "religion", y = "millions of followers", 
         title = "Most Popular Religions",
         subtitle = "[2021]",
         caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") +
    theme(plot.title = element_text(size = 18, color = "blue"),
        plot.subtitle = element_text(size = 14, color = "gold")) +
          coord_flip()
```

# Haig Bedros' Expansion:  

# Exploring Cumulative Distribution with tidyverse

## 3A: Cumulative Followers Distribution
**Objective:** We aim to understand the cumulative distribution of religious followers. This will provide insights into how a few religions may dominate in terms of followership.

**Approach:** We'll sort the religions by the number of followers and calculate the cumulative sum of followers. Then, we'll visualize this data to see the progression.

**Key Functions Used:**

- `arrange():` Orders rows by specific columns.
- `mutate()`: Used to add new columns or modify existing ones.
- `cumsum()`: Computes the cumulative sum of a numeric column.

```{r}
# Calculate cumulative followers for each religion
religions_cumulative <- religions %>%
  arrange(desc(millions_of_followers)) %>%
  mutate(cumulative_followers = cumsum(millions_of_followers))

# Plot the cumulative distribution
ggplot(religions_cumulative, aes(x = reorder(religion, -cumulative_followers), y = cumulative_followers)) + 
  geom_col(fill = "lightblue") + 
  labs(x = "Religion", y = "Cumulative Followers (in millions)", 
       title = "Cumulative Distribution of Religious Followers", 
       caption = "https://en.wikipedia.org/wiki/List_of_religious_populations") + 
  coord_flip()

```

## 3B: Distribution of Top Religions' Followers  
**Objective:** Let's delve deeper into the top 5 religions and understand the distribution of their followers.  

**Approach:** We'll use a box plot, which can show the central tendency and spread of the followers for these top religions.  

**Key Functions Used:**  

- `filter():` Filters rows based on certain conditions.
- `boxplot():` Creates a box plot for visualizing the distribution of a numeric variable.

```{r}
# Filtering the top 5 religions
top5_religions <- religions %>%
  filter(religion %in% top5)

# Plotting the distribution of followers for the top 5 religions
ggplot(top5_religions, aes(x = religion, y = millions_of_followers)) + 
  geom_boxplot(fill = "lightblue") + 
  labs(x = "Religion", y = "Followers (in millions)", 
       title = "Distribution of Followers for Top 5 Religions", 
       caption = "https://en.wikipedia.org/wiki/List_of_religious_populations")
```

By expanding the analysis, we can derive richer insights from the dataset and understand the distribution of religious followers more comprehensively. The use of functions like `cumsum()` and `geom_boxplot()` allows us to explore the data from different angles, showcasing the versatility of the `tidyverse` ecosystem.
